---
title: "DBA3702 Assignment 2"
output: html_document
date: "26 August 2021"

---
# Question 1

```{r}

# Remove all rows with empty second cols
raw.data <- read.csv("./Univ Education.csv", stringsAsFactors = F)
colnames(raw.data) <- 1:ncol(raw.data)
data.clean <- raw.data[raw.data[[2]] != "" & raw.data[[2]] != " " & (grepl("Males:", raw.data$"1") | grepl("Females:", raw.data$"1") | grepl("Variables", raw.data$"1")),]

# Set new header
names(data.clean) <- data.clean[1,]
data.clean <- data.clean[-1,]
# To validate if data is appropriately cleaned
# nrow(data.clean)

print(data.clean)

```
# Question 2

```{r}
library("stringr")
library("dplyr")
library("tidyr")

# Add Gender and Major columns to data frame
data.clean$Gender <- ifelse(grepl("Males", data.clean$` Variables `), "Males", ifelse(grepl("Females", data.clean$` Variables `), "Females", "na"))

data.clean$Major <- str_remove(data.clean$` Variables `, "Males: ")
data.clean$Major <- str_remove(data.clean$Major, "Females: ")
data.clean$Major <- str_trim(data.clean$Major)

data.clean <- gather(data.clean, "Year", "Count", -` Variables `,-Gender, -Major)

df <- data.frame(Year = data.clean$Year, Gender = data.clean$Gender, Major = data.clean$Major, Count = data.clean$Count)

df$Count <- gsub(",","",df$Count)
df$Count <- gsub("na",0,df$Count)
df <- transform(df, Count = as.integer(Count))
dfWithoutOthers <- df

# Accounting for graduates whose major cannot be classified in any of the major categories provided in 2017
maleTotal_2017 <-  as.integer(gsub(",","", raw.data[5,26]))
femaleTotal_2017 <-  as.integer(gsub(",","", raw.data[21,26]))
prelimAggregate <- aggregate(Count~Gender+Year, df, sum)
maleOther2017 <- maleTotal_2017 - prelimAggregate[50, "Count"]
femaleOther2017 <- femaleTotal_2017 - prelimAggregate[49, "Count"]

# Adding these graduates under the "Others" category for the year 2017
df <- add_row(df, Year = "2017", Gender = "Males", Major = "Others", Count = maleOther2017)
df <- add_row(df, Year = "2017", Gender = "Females", Major = "Others", Count = femaleOther2017)

print(df)
```

# Question 3

***There is a higher number of female graduates than male graduates in total across all majors from 1993-2017***

- Total number of male graduates: 140,794
- Total number of female graduates: 146,150

**For data excluding 2017 graduates whose main field of study is not classifiable under one of the majors**

- Total number of male graduates: 140,783
- Total number of female graduates: 146,115

```{r}
# Inclusive of all graduates
aggregate(Count~Gender, df, sum)
# Excluding 2017 graduates whose main field of study is not classifiable under one of the majors
aggregate(Count~Gender, dfWithoutOthers, sum)

```
# Question 4

***2017 has the largest number of dentistry graduates***
```{r}

dentistry <- aggregate(Count~Major+Year, df, sum)
dentistry <- dentistry %>% subset(Major == "Dentistry")
dentistry[which.max(dentistry$Count),]

```
# Question 5

***Most favourable major for Females from 1993-2017:***<br>

- 1993 - 2003 -> Humanities & Social Sciences
- 2004 - 2008 -> Engineering Sciences
- 2009 - 2017 -> Humanities & Social Sciences

```{r}

trend <- subset(df, Gender =="Females") %>% group_by(Year) %>% top_n(1, Count)
print(trend)
```




