---
title: "DBA3702 Assignment 4"
output: html_document
date: "10 Sept 2021"
---
# Question 1



# Question 2

### Function to validate if input sequence is in the right format

```{r}
# Based on the format covered during our lecture, a sequence of the correct format should
# 1. Only contain "F" "M" and "," characters
# 2. Should not contain any white-space
# 3. Should be delimited by ","
# 4. Should have odd number of characters (derived constraint)
seqChecker <- function(seq) {
  
  # Check if sequence has an odd number of characters --> return false if even
  if(nchar(seq) %% 2 == 0) {
    return(FALSE)
  }
  # Check if last character is not "M" or "F"
  if(substr(seq, nchar(seq),nchar(seq)) != "F" & substr(seq, nchar(seq),nchar(seq)) != "M") {
    return(FALSE)
  }
  # Split the seq by its delimiter
  seq_split <- strsplit(seq, ",")[[1]]
  if(length(seq_split < 3)) {
    return(FALSE)
  }
  
  # Check if split vector contains any element besides M and F
  if(length(seq_split[seq_split %in% c("F", "M")]) != length(seq_split)) {
    return(FALSE)
  }
  return(TRUE)
}
```

### Functions handlers for all error patterns identified in the column.
#### The following cases should be considered:
1. Differences in Capitalization
1. F / M represented as Female / Male
1. Extra white-spaces btw delimiters
1. Invalid characters (i.e. <)
1. Empty Sequences
1. Wrong spelling
1. Missing delimiters

```{r}
# Handler to capitalize all chars in the sequence provided
capsHandler <- function(seq) {
  return(toupper(seq))
}
# Handler to replace all instances of Female/Male with shorthands F/M
shortformHandler <- function(seq) {
  result <- gsub("MALE", "M", seq)
  return(gsub("FEM", "F", result))
}

# Handler to remove all chars not upper case letters or isn't a comma (spaces are also removed)
specialCharHandler <- function(seq) {
  return(gsub("[^[:upper:]\\,]", "", seq))
}

# Handler to recombine vector of chars into string
strRecombineHandler <- function(seqVector) {
  stringStore <- ""
  needComma <- FALSE
  for(char in seqVector) {
    if(!needComma && char != ",") {
      stringStore <- paste(stringStore, char, sep = "")
      needComma <- TRUE  
    } else if(needComma && char == ",") {
        stringStore <- paste(stringStore, char, sep = "")
        needComma <- FALSE 
    } else if(!needComma && char == ",") {
        # Ignore the extra comma
    } else {
        stringStore <- paste(stringStore, ",", char, sep ="")
        needComma <- TRUE
    }
  }
  return(stringStore)
}
# Handler for special cases such as spelling errors
spellingErrorHandler <- function(seq){
  if(seq == "famale") {
    return("F")
  }
}

```

Now, it's time to put all the handlers into action and clean the column:
**(Rows with less than 3 data points in the gender sequence are removed as they are insufficient for analysis according to prof during lecture)**

```{r}

raw.data <- read.csv("The family with the largest number of children.csv", fileEncoding = "UTF-8-BOM")
clean.data <- raw.data
emptyRows <- c() 

for(rowNum in 1:nrow(clean.data)) {
  seq <- clean.data[rowNum, "Gender.Sequence"]
  # Check if gender sequence is empty
  if (seq == "") {
    emptyRows <- c(emptyRows, rowNum)
  } else {
      # If sequence fails validation, handle with handlers
      if(!seqChecker(seq)) {
        tempSeq <- capsHandler(seq)
        tempSeq <- shortformHandler(tempSeq)
        tempSeq <- specialCharHandler(tempSeq)
        tempSeqSplit <- strsplit(tempSeq, "")[[1]]
        if(length(tempSeqSplit[tempSeqSplit %in% c("F", "M", ",")]) != length(tempSeqSplit)) {
          tempSeq <- spellingErrorHandler(seq)
        } else {
          tempSeq <- strRecombineHandler(tempSeqSplit)
        }
        # Removes rows with gender sequence of less than 3 data points
        if(nchar(tempSeq) < 5) {
            emptyRows <- c(emptyRows, rowNum)
        }
        clean.data[rowNum, "Gender.Sequence"] <- tempSeq
      }
  }
}
# Rows with empty gender seq are removed (eg. row 76 in raw data)
clean.data <- clean.data[-emptyRows,]
clean.data
```